\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-12 paper.tex

\section*{Locally unary nodes}
The arity of a node is the number of child nodes it has. There is an
important distinction between the number of children a node has
in the graph and the number of children it has in the local trees,
because a (graph) child node is only a child in the local tree for position
$x$ if $x$ is in the inheritance interval for the corresponding edge.
Thus, for a given node $u$ and a genome position $x$ the arity
of $u$ in that local tree is less than or equal to the graph arity of
$u$. The arity
of a node in the local trees can therefore change as we move along
the genome.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence unary nodes (see previous section),
which we illustrate by an example simulation.
[THIS SIMULATION...]

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent
(legend in Fig.~\ref{fig-inferred-args})
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes where they are locally unary.
}
\end{figure}

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.

In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference results for
(A) \kwarg, (B) \argweaver, (C) \tsinfer,
and (D) \relate\
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}.
The \argweaver\ example
is a sample from the inferred posterior distribution.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Line width corresponds to the genomic span of the edge.
Position on the X and Y axes for the graph visualization is arbitrary.
 Node colour
refers to the number of parents of the node.
The fill colour intensity of a node is determined by the genomic span over which
it has two or more children divided by the span where it has any
children (the ``coalescent span'').
The genomic intervals associated with edges are shown below
each ARG, where each row is an edge
and the X axis reflects genomic position.
Vertical lines indicate breakpoints between local trees.
}
\end{figure}

>>>>

\section*{Locally unary nodes}
We can define the arity of a node as the number of child nodes it has.
There is an
important distinction between the number of children a node has
in the graph and the number of children it has in the local trees,
because a (graph) child node is only a child in the local tree for position
$x$ if $x$ is in the inheritance interval for the corresponding edge.
Thus, for a given node $u$ and a genome position $x$ the arity
of $u$ in that local tree is less than or equal to the graph arity of
$u$. The arity
of a node in the local trees can therefore change as we move along
the genome.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence unary nodes (see previous section),
which we illustrate by an example simulated ARG, which
undergoes a series of successive steps of simplification
in Fig.~\ref{fig-simplification}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
(A) An example ARG simulated from a diploid Wright-Fisher model.
(B) Remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes in local trees in which they are unary.
In each case, the graph is shown on the left
and corresponding local trees on the right.
Graph nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent;
see the text for more details.
}
\end{figure}

The ARG in Fig.~\ref{fig-simplification}A is the output of a
backwards-time Wright-Fisher simulation for a sample of 2 diploid
individuals (population size $N=10$), and follows a similar process
to the methods described by~\cite{nelson2020accounting}.
As we proceed backwards in time, generation by generation, the
extant lineages chose parents randomly.
With a certain probability recombination occurs, and the ancestral
material of a lineage is split between the two parental
genomes. Local coalesence occurs
when lineages with overlapping ancestral material choose the same
parent genome.
Note that in this simulation we do not explicitly
model recombination \emph{events} via an ARG node, but simply record
the \emph{effects} of a recombination via outbound edges to
the parent individual's two genomes.
Thus, a recombinant node may also correspond to a coalescence
(e.g.~node \noderef{g} in Fig.~\ref{fig-simplification}).
The distinction of
using a single node to represent
a recombination event (e.g.~Fig~\ref{fig-ancestry-resolution}),
or two to represent the parent genomes is usually not important
(either is possible in the gARG encoding)
and the most convenient approach will vary by application.
Note also that polytomies are a natural feature of such
a Wright-Fisher model
(e.g.~node \noderef{k} in Fig.~\ref{fig-simplification}).

Fig.~\ref{fig-simplification} shows four ARGs (A--D), with
the graph depicted on the left and corresponding sequence
of local trees on the right. The graph visualisations have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the size of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinanants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?

Returning to the main topic of this section,
Fig.~\ref{fig-simplification}A is the direct output of a Wright-Fisher
simulation in which we retain all nodes involved in recombination
or common ancestry events. This is the true history, and contains
a very high level of detail, some of which may be considered
redundant (or, from another perspective, unobservable).
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference results for
(A) \kwarg, (B) \argweaver, (C) \tsinfer,
and (D) \relate\
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}.
The \argweaver\ example
is a sample from the inferred posterior distribution.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Position on the X and Y axes for the graph visualization is arbitrary.
Line width and node colour are as described in Fig.\ref{fig-simplification}.
The genomic intervals associated with edges are shown below
each ARG, where each row is an edge
and the X axis reflects genomic position.
Vertical lines indicate breakpoints between local trees.
}
\end{figure}
